Galaxy Imaging with Generative Models: Insights from a Two-Models Framework

### 2025-03-16 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS_r1.tex
rename files with r1 tag

However, Figure~\ref{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4} presents the results of the "discriminator test" for models trained with datasets of sizes $N=10^4$ (panel $(a)$) and $N=10^3$ (panel $(b)$), evaluated at \textit{epoch} 100. Readers are encouraged to compare these findings with the corresponding panel in Figure~\ref{fig-gan-hinge-loss-evol}. When training with $N=10^4$ or $N=10^3$, the results indicate that the discriminators are not entirely agnostic to the origin of the image samples, as evidenced by the distinct distributions of $\ell(D_A(\bm{x}_A), D_A(\bm{x}_B))$ and \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_{\text{train}}))\) (and similarly for the $D_B$ discriminator). 
Moreover, the discrepancies in these loss distributions persist even when training continues beyond \textit{epoch} 100, suggesting that the observed divergences are not merely attributable to early convergence issues. It could be argued that these inconsistencies may arise from suboptimal hyperparameter choices in designing the \texttt{light-weight-gan} architecture used for the experiment, rather than solely from the limited dataset size. Notably, in the absence of convergence across the four loss distributions, there is no \textit{a priori} method to guide further investigations or remedial measures effectively.
%
\begin{figure}
    \centering
	\includegraphics[width=0.7\linewidth]{fig-gini-m20-gan-compare_0.pdf}
	\caption{Some morphological coefficients (see Figure~\ref{fig-morpho-coeff}) computed using \texttt{light-weight-gan} models trained with different dataset sizes ($N=10^5, 10^4, 10^3$).}
	\label{fig-morpho-coeff-gan-comparison}
\end{figure}
\begin{figure}
    \centering
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_10000_A_B_num_100.pdf}
        \caption{$N=10^4$}
        \end{subfigure}%
    \hfill
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_1000_A_B_num_100.pdf}
        \caption{$N=10^3$}
        \end{subfigure}%
	\caption{Same legend as for Figure~\ref{fig-gan-hinge-loss-evol} but with different dataset sizes: $N=10^4$ (a) and $N=10^3$ (b).}
	\label{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4}
\end{figure}
%
\section{Conclusion}
\label{sec-conclusion}
%
In this study, we investigated the behavior of generative models applied to galaxy image generation, focusing on their performance under controlled experimental conditions. We compared three architectures: the \texttt{light-weight-gan} GAN-based model, the \texttt{Glow} flow-based model, and the diffusion model based on a \texttt{U-Net} denoiser (Section~\ref{sec-Exp-Models}). These models were trained and evaluated using identical datasets sourced from the SDSS {\color{red}DR9} survey (Section~\ref{sec-Dataset}).

@@

In conclusion, while modern generative models can produce high-quality galaxy images with reasonably sized datasets, much remains to be explored regarding the mechanisms governing their learning processes and the factors influencing their generalization. The "two-models" methodology proved to be a valuable approach, allowing us to assess both the similarity of generated images to real-world data and the consistency of the learned probability distributions. This framework represents a key takeaway of this study to study generative modelling. The methods developed here are certainly not limited to generating images of galaxies.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
We thank Prafulla Dhariwal who have accepted the reproduction of Figure 2 of  \cite{Kingma2018} (see Figure~\ref{fig-Glow-archi}). We thank Kim Seonghyeon for fruitful discussion on \texttt{Glow} model implementation. We acknowledge the use of  Nvidia’s V100 resources from French GENCI–IDRIS (Grant 2024-AD010413957R1).

>>>>

However, Figure~\ref{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4} presents the results of the "discriminator test" for models trained with datasets of sizes $N=10^4$ (panel $(a)$) and $N=10^3$ (panel $(b)$), evaluated at \textit{epoch} 100. Readers are encouraged to compare these findings with the corresponding panel in Figure~\ref{fig-gan-hinge-loss-evol}. When training with $N=10^4$ or $N=10^3$, the results indicate that the discriminators are not entirely agnostic to the origin of the image samples, as evidenced by the distinct distributions of $\ell(D_A(\bm{x}_A), D_A(\bm{x}_B))$ and \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_{\text{train}}))\) (and similarly for the $D_B$ discriminator). 
Moreover, the discrepancies in these loss distributions persist even when training continues beyond \textit{epoch} 100, suggesting that the observed divergences are not merely attributable to early convergence issues. It could be argued that these inconsistencies may arise from suboptimal hyperparameter choices in designing the \texttt{light-weight-gan} architecture used for the experiment, rather than solely from the limited dataset size. Notably, in the absence of convergence across the four loss distributions, there is no \textit{a priori} method to guide further investigations or remedial measures effectively.
%
\begin{figure}
    \centering
	\includegraphics[width=0.7\linewidth]{fig-gini-m20-gan-compare_0.pdf}
	\caption{Some morphological coefficients (see Figure~\ref{fig-morpho-coeff}) computed using \texttt{light-weight-gan} models trained with different dataset sizes ($N=10^5, 10^4, 10^3$).}
	\label{fig-morpho-coeff-gan-comparison}
\end{figure}
\begin{figure}
    \centering
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_10000_A_B_num_100.pdf}
        \caption{$N=10^4$}
        \end{subfigure}%
    \hfill
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_1000_A_B_num_100.pdf}
        \caption{$N=10^3$}
        \end{subfigure}%
	\caption{Same legend as for Figure~\ref{fig-gan-hinge-loss-evol} but with different dataset sizes: $N=10^4$ (a) and $N=10^3$ (b).}
	\label{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4}
\end{figure}
%
\section{Conclusion}
\label{sec-conclusion}
%
In this study, we investigated the behavior of generative models applied to galaxy image generation, focusing on their performance under controlled experimental conditions. We {\color{red} have presented three types of models as generic examples}: the \texttt{light-weight-gan} GAN-based model, the \texttt{Glow} flow-based model, and the diffusion model based on a \texttt{U-Net} denoiser (Section~\ref{sec-Exp-Models}). These models were trained and evaluated using identical datasets sourced from the SDSS {\color{red}DR9} survey (Section~\ref{sec-Dataset}).

@@

In conclusion, while modern generative models can produce high-quality galaxy images with reasonably sized datasets, much remains to be explored regarding the mechanisms governing their learning processes and the factors influencing their generalization.  
{\color{red}
As a key takeaway, we first highlight the necessity of sufficiently large training datasets to ensure effective learning. The exact amount of data required is not yet settled and remains an open research question, but it certainly depends on the image resolution, the number of model parameters, and the model architecture. Investigating different strategies to improve generalization without requiring massive datasets remains an open question. In future work, one could explore restricting the diversity of galaxies in the dataset or leveraging pre-trained models on general-purpose images to enhance training efficiency.  
Secondly, we emphasize the importance of evaluations beyond morphological distributions to properly assess whether model optimization is effective. The "two-models" methodology proved to be a valuable approach, enabling us to assess both the similarity of generated images to real-world data and the consistency of the learned probability distributions. This framework has been a key element of our study.  
Finally, we argue that score-based diffusion models offer a direct way to test whether they operate in the generalization regime.

We believe that the methods developed here are not limited to generating images of galaxies or to the specific models we have used.  
}

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
We thank Prafulla Dhariwal who have accepted the reproduction of Figure 2 of  \cite{Kingma2018} (see Figure~\ref{fig-Glow-archi}). We thank Kim Seonghyeon for fruitful discussion on \texttt{Glow} model implementation. We acknowledge the use of  Nvidia’s V100 resources from French GENCI–IDRIS (Grant 2024-AD010413957R1).
